---
title: "Getting started with omicsaudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with omicsaudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 5.5, fig.height = 3.6)
library(omicsaudit)
```

## The idea

A *pipeline* is a function `pipeline(X, y)`. `X` is a raw samples x features
matrix (linear scale, `NA` for missing) and `y` a 0/1 outcome. It returns
`list(selected = <character>, performance = <number>)`. omicsaudit never reads
the pipeline's code; it controls what goes in and watches what comes out.

## Simulated data with a known answer

```{r sim}
sim <- simulate_omics(n = 80, p = 150, n_signal = 10, effect = 1.5,
                      batch_effect = 0.5, missing_rate = 0.2, seed = 1)
sim
```

The signal (`sim$truth`), the batch structure and an abundance-dependent
(MNAR) missingness pattern are all known.

## Run a pipeline and audit it

```{r audit}
pipe <- zoo_load("reference_nested_cv")$pipeline
pipe(sim$X, sim$y)$performance

a <- audit_standard(pipe, sim$X, sim$y, batch = sim$batch, quick = TRUE, seed = 1)
a
```

The standard audit runs a shuffled-label test through the whole pipeline
(`audit_null`), a batch-confounding check (`audit_confound`) and a feature
stability check (`audit_stability`, reported but not part of the verdict).

## How small an effect can this pipeline find?

`audit_planted()` scrambles the outcome, plants a known fold change into random
features and records how many of them the pipeline recovers.

```{r planted}
top5 <- function(X, y) {
  L <- log2(X); L[is.na(L)] <- min(L, na.rm = TRUE)
  t <- abs(apply(L, 2, function(v) stats::t.test(v ~ y)$statistic))
  list(selected = names(sort(t, decreasing = TRUE))[1:5], performance = NA_real_)
}
p <- audit_planted(top5, sim$X, sim$y, effects = c(0, 0.5, 1, 1.5, 2),
                   n_plant = 3, reps = 10, n_grid = c(40, 80), seed = 1)
p
plot(p)
samples_needed(p, effect = 0.25)
```

### Planting and MNAR missingness: approach and limits

* Planting multiplies the **observed** values of the chosen features in class 1
  by `2^effect`. Missing values stay missing and nothing new is masked, so the
  missingness pattern is unchanged (this is tested).
* In real MNAR data a higher true abundance is also less likely to be missing.
  Planting cannot undo values that were already missing, so planted features keep
  slightly more missingness in class 1 than a truly higher-abundance feature
  would. For positive effects this makes the audit conservative.
* Only upward shifts in class 1 are planted. A downward shift would need
  synthetic extra dropout based on an assumed MNAR curve.
* `samples_needed()` extrapolates the MDE with a power law in n. It is reliable
  close to the sample sizes tested and should be treated as a rough guide beyond.

## Which analysis decision matters?

```{r variants}
g <- variant_grid(variant_pipeline, norm = c("none", "median"),
                  impute = c("min", "median"), max_missing = c(0.5, 1))
v <- audit_variants(g$variants, sim$X, sim$y, design = g$design)
v
```

## The zoo

```{r zoo}
zoo_list()
zoo_load("flawed_select_before_split")$meta$description
```

`zoo_score(audit_fn)` generates each entry's dataset, runs `audit_fn` and
reports the detection rate on the flawed pipelines and the false-alarm rate on
the references. It takes a few minutes with the default settings, so it is not
run here:

```r
zoo_score(audit_standard, quick = FALSE, n_perm = 30)
```

Any function `f(pipeline, X, y, batch, groups, quick, seed, ...)` returning an
object with a logical `flagged` can be benchmarked the same way.

## A one-page report

```{r report, eval = FALSE}
audit_report(a, planted = p, variants = v, file = "audit.html",
             pipeline_name = "reference_nested_cv")
```

## Not yet covered

omicsaudit does not yet call `bioLeak` or `nestedcv`; overlap with them has been
noted but not integrated, and no novelty claim is made until a literature and
package search has been done.
